---
title: "R Querying and Viz"
author: "Sofia Sarak"
date: "May 15, 2026"
format: html
editor: visual
---

The following notebook contains SQL queries and a visualization based in answering the question: **do wealthier countries have healthier cities**? UN Urban Indicators and World Bank Group data is used to make this comparison.

## Setup

```{r}
#| warning: false
#| message: false
# load necessary libraries
library(tidyverse)
library(DBI)
library(duckdb)
library(dplyr)
library(patchwork)
library(showtext)
```

```{r}
# connect to database using duckdb
conn <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = "data/public-spaces")
# check connection
conn
```

```{r}
# view contents of database
DBI::dbListTables(conn)
```

## SQL Queries

Here, I run three separate SQL queries, as my final visualization will feature three different urban resource variables:

-   average share of green area in a city

-   the percent of a city's population that has easy access to public open space

-   the percent of a city's population that has easy access to public transportation

A more detailed description can be found on the UN Urban Indicator Database's website (see repository README).

### Average share of green area

```{r}
green_areas_avg <- dbGetQuery(conn, "SELECT AVG(green_area_share) AS avg_green_area, 
ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT year, green_area_share, city_name, country_or_territory_code FROM 
    Green_Areas G JOIN City C USING (city_code)) J USING (country_or_territory_code)
    WHERE year = 2020
    GROUP BY country_or_territory_name")
```

### Public open space access

```{r}
open_space_pop_avg <- dbGetQuery(conn, "SELECT AVG(open_space_pop) AS avg_open_pop, 
ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT reference_year, open_space_pop, city_name, country_or_territory_code FROM 
    Open_space G JOIN City C USING (city_code)) J USING (country_or_territory_code)
    WHERE reference_year = 2020
    GROUP BY country_or_territory_name")
```

### Public transportation access

```{r}
transit_avg <- dbGetQuery(conn, "SELECT AVG(transport_pop) AS avg_transport_pop, 
ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT reference_year, transport_pop, city_name, country_or_territory_code FROM 
    Transport G JOIN City C USING (city_code)) J USING (country_or_territory_code)
    WHERE reference_year = 2020
    GROUP BY country_or_territory_name")
```

Each query contains two separate joins: the most inner links the urban resource table (Green_Areas, Open_space, or Transport) to the City table on city code, while the outer joins those two tables to Country table (which houses GDP information). 

The queries also aggregate the urban resource variable to its mean by country (since there are multiple cities with data in each country). 

## Prepare data for visualization

The three dataframes differ in which countries the data is available for. I want the number of countries to be consistent across all three plots, so I will inner join all three on country to keep only those that match everywhere.

```{r}
#| message: false
#| warning: false

# inner join the three data frames on country
join <-  green_areas_avg %>% 
  inner_join(open_space_pop_avg, by = "country") %>% 
  inner_join(transit_avg, by = "country")
# this data frame contains all of the matching countries in our three datasets

# filter each original df to match this list of countries
green_areas_avg <- green_areas_avg %>% 
  filter(country %in% join$country)

open_space_pop_avg <- open_space_pop_avg %>% 
  filter(country %in% join$country)

transit_avg <- transit_avg %>% 
  filter(country %in% join$country)
```

There are a few data points with really high GDP. I filter them out before plotting, since including them in the visualization smushes the rest of the data, making the relationship between variables difficult to discern.

```{r}
# filter three data frames by GDP
green_areas_avg <- green_areas_avg %>% 
  filter(gdp < 10000000000000)

transit_avg <- transit_avg %>% 
  filter(gdp < 10000000000000)

open_space_pop_avg <- open_space_pop_avg %>% 
  filter(gdp < 10000000000000)
```

## Prepare theme elements

```{r}
# save palette colors
mediumgreen <-  "#606C38"
darkgreen <- "#283618"
bg <- "#FEFAE0"
mediumorange <- "#DDA15E"
darkorange <- "#BC6C25"

# load fonts
font_add_google(name = "Montserrat", family = "mont")
font_add_google(name = "Google Sans", family = "google")
```

```{r}
# create ggplot theme to apply to all 3 plots
# the majority of this code changes the colors and sizes of plot elements
my_theme <- function(base_size = 17){
  theme_minimal(base_size = base_size) +
    theme(
    panel.grid = element_line(color = mediumorange, linewidth = 0.2),
    plot.title = element_text(family = "mont",
                                  size = rel(0.99),
                                  margin = margin(b = 7)),
    plot.background = element_rect(fill = bg,
                                   color = bg),
    panel.background = element_rect(fill = bg, 
                                    color = bg),
        plot.subtitle = element_text(family = "google", size = rel(0.70)),
        axis.text.y = element_text(hjust = 0, family = "google", size = rel(0.5), color = darkorange),
        axis.text.x = element_text(family = "google", size = rel(0.5), color = darkorange),
        axis.title = element_text(family = "google", size = rel(0.6), color = darkorange),
    )
}
```

## Plot

Here, I create three separate scatterplots for each urban resource variable, then join them all using the {patchwork} package.

```{r}
#| warning: false
#| message: false
showtext_auto(enable = TRUE)

green <- ggplot(green_areas_avg, aes(avg_green_area, gdp)) +
  geom_point(alpha = 0.5,
             color = mediumgreen,
             fill = mediumgreen) +
  geom_smooth(method = "lm", # calculate linear relationship
              se = FALSE, # remove confidence interval
              linetype = 2, # dotted line
              linewidth = 1,
              color = darkgreen) + 
  labs(x = "Average % Share\nof Green Area",
       y = "National GDP (USD 2020)") +
  
  # adjust y-axis labels to read as "trillion"
  # set y limits to start at 0 -- geom_smooth goes below 0, throwing off axis
  scale_y_continuous( labels = scales::label_number(scale = 1e-12, suffix = " trillion"),
                      limits = c(0, NA)) +
  my_theme()

open <- ggplot(open_space_pop_avg, aes(avg_open_pop, gdp)) +
  geom_point(alpha = 0.5,
             color = mediumgreen,
             fill = mediumgreen) +
  geom_smooth(method = "lm",
              se = FALSE,
              linetype = 2, # dotted line
              linewidth = 1,
              color = darkgreen) +
   labs(x = "Average % of Population\nwith Access to Open Space",
       y = "National GDP (USD 2020)") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-12, suffix = " trillion"),
                     limits = c(0, NA)) +
  my_theme()

transit <- ggplot(transit_avg, aes(avg_transport_pop, gdp)) +
  geom_point(alpha = 0.5,
             color = mediumgreen,
             fill = mediumgreen) +
  geom_smooth(method = "lm",
              se = FALSE,
              linetype = 2, # dotted line
              linewidth = 1,
              color = darkgreen) +
   labs(x = "Average % of Population\nwith Access to Public Transit",
       y = "National GDP (USD 2020)") +
   scale_y_continuous(labels = scales::label_number(scale = 1e-12, suffix = " trillion"),
                      limits = c(0, NA))+
  my_theme()

# join 3 plots into 1
green + open + transit +
  plot_layout(axes="collect") + # specify shared y-axis
  
  # plot title, subtitle, and caption, along with their associated theming
  plot_annotation(title = "Urban Resources & GDP: 111 countries worldwide",
                  subtitle = "Greater economic prosperity correlates with greater resources, such as\ngreen area, open space, and public transportation.",
                  caption = 'Access is defined as "Share of urban population with convenient access to public transport" by the United Nations.',
                  theme = theme(plot.background = element_rect(fill  = bg),
                                plot.title = element_text(family = "mont",
                                                          face = "bold",
                                  size = 17,
                                  margin = margin(t = 7, b = 7, r = 7), # adjusts spacing around title
                                  color = darkgreen),
                                plot.subtitle = element_text(family = "mont",
                                  size = 12,
                                  margin = margin(b = 7),
                                  color = darkgreen),
                                plot.caption = element_text(family = "mont",
                                  size = 8,
                                 # margin = margin(b = 7),
                                  color = darkgreen)))

# ggsave("scatterplot.png")

showtext_auto(enable = FALSE)
```

## Closing the connection!!

```{r}
# close duckdb connection
DBI::dbDisconnect(conn, shutdown = TRUE)
```
