---
title: "R Querying and Viz"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(DBI)
library(duckdb)
library(dplyr)
library(patchwork)
library(showtext)
```
## Connecting to a database
```{r}
conn <- DBI::dbConnect(duckdb::duckdb(),
                       dbdir = "data/public-spaces")

conn
```


```{r}
DBI::dbListTables(conn)
```

```{r}
open <- tbl(conn, "Open_space")
```
```{r}
transport <- tbl(conn, "Transport")
```



```{r}
transport
```

```{r}
green
```

```{r}
open
```


```{r}
green <- tbl(conn, "Green_Areas")
country <- tbl(conn, "Country")
city <- tbl(conn, "City")
```

```{r}
join_areas <-  open %>% 
  left_join(green) %>% 
  select(city_code, open_space_area_share, green_area_share) %>% 
  collect()
```
```{r}
join_country <- join_areas %>% 
  left_join(city, copy = TRUE) %>% 
  collect()
```

Create city country join
```{r}
loc <- dbGetQuery(conn, "SELECT city_code, city_name, country_or_territory_code, country_or_territory_name, annual_gdp FROM
    Country 
    JOIN City USING (country_or_territory_code)")

# view
loc
```

Join location to joined open and green areas df and average
```{r}
final <- join_areas %>% 
  left_join(loc) %>% 
  group_by(country_or_territory_name) %>% 
  summarize(avg_green_area_share = mean(green_area_share, na.rm = TRUE),
            avg_open_space_share = mean(open_space_area_share, na.rm = TRUE)) %>% 
  collect()
```

plot
```{r}
ggplot(final, aes(avg_green_area_share, avg_open_space_share)) +
  geom_point() +
  geom_smooth(method = "lm")
```


```{r}
green_areas <- dbGetQuery(conn, "SELECT AVG(green_area_share), AVG(open_space_area_share) FROM Open_space O JOIN
(SELECT year, green_area_share, city_name, country_or_territory_code, country_or_territory_name, annual_gdp FROM
    Country C JOIN 
    (SELECT year, green_area_share, city_name FROM Green_Areas G 
    JOIN City C USING (city_code)) 
    J USING (country_or_territory_code)) USING (country_or_territory_code)
    WHERE year = 2020")
```

```{r}
green_areas_avg <- dbGetQuery(conn, "SELECT AVG(green_area_share) AS avg_green_area, 
ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT year, green_area_share, city_name, country_or_territory_code FROM 
    Green_Areas G JOIN City C USING (city_code)) J USING (country_or_territory_code)
     WHERE year = 2020
    GROUP BY country_or_territory_name")
```

```{r}
open_space_avg <- dbGetQuery(conn, "SELECT AVG(open_space_share) AS avg_open_share, 
ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT reference_year, open_space_area_share, city_name, country_or_territory_code FROM 
    Open_space G JOIN City C USING (city_code)) J USING (country_or_territory_code)
     WHERE reference_year = 2020
    GROUP BY country_or_territory_name")
```


```{r}
open_space_pop_avg <- dbGetQuery(conn, "SELECT AVG(open_space_pop) AS avg_open_pop, 
ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT reference_year, open_space_pop, city_name, country_or_territory_code FROM 
    Open_space G JOIN City C USING (city_code)) J USING (country_or_territory_code)
     WHERE reference_year = 2020
    GROUP BY country_or_territory_name")
```



```{r}
transit_avg <- dbGetQuery(conn, "SELECT AVG(transport_pop) AS avg_transport_pop, 
ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT reference_year, transport_pop, city_name, country_or_territory_code FROM 
    Transport G JOIN City C USING (city_code)) J USING (country_or_territory_code)
     WHERE reference_year = 2020
    GROUP BY country_or_territory_name")
```


```{r}
green_areas_avg_top10 <- dbGetQuery(conn, "SELECT AVG(green_area_share) AS avg_green_area, ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT year, green_area_share, city_name, country_or_territory_code FROM 
    Green_Areas G JOIN City C USING (city_code)) J USING (country_or_territory_code)
    WHERE year = 2020
    GROUP BY country_or_territory_name
    ORDER BY ANY_VALUE(annual_gdp) DESC
    LIMIT 10")
```

```{r}
green_areas_avg_bot10 <- dbGetQuery(conn, "SELECT AVG(green_area_share) AS avg_green_area, ANY_VALUE(country_or_territory_name) AS country, ANY_VALUE(annual_gdp) AS gdp FROM
    Country C 
    JOIN (SELECT year, green_area_share, city_name, country_or_territory_code FROM 
    Green_Areas G JOIN City C USING (city_code)) J USING (country_or_territory_code)
    WHERE year = 2020
    GROUP BY country_or_territory_name
    ORDER BY ANY_VALUE(annual_gdp)
    LIMIT 10")
```

```{r}
green_open <- dbGetQuery(conn, "SELECT AVG(green_area_share) AS avg_green_area, ANY_VALUE(country_or_territory_name) AS country, AVG() AS gdp FROM
    Country C 
    JOIN (SELECT year, green_area_share, city_name, country_or_territory_code FROM 
    Green_Areas G JOIN City C USING (city_code)) J USING (country_or_territory_code)
    WHERE year = 2020
    GROUP BY country_or_territory_name
    ORDER BY ANY_VALUE(annual_gdp)
    LIMIT 10")
```



```{r}
green_areas_avg_top10
```


```{r}
green_areas_avg
```
```{r}
open_space_avg
```

## Prep data for visualization

The three dataframes differ in which countries the data is available for. I want the number of countries to be consistent across all three plots, so I will inner join all three on country to keep only those that match everywhere.
```{r}
#| message: false
#| warning: false
join <-  green_areas_avg %>% 
  inner_join(open_space_pop_avg, by = "country") %>% 
  inner_join(transit_avg, by = "country")

# this data frame contains all of the matching countries in our three datasets
# we will filter each original df to match this list of countries
green_areas_avg <- green_areas_avg %>% 
  filter(country %in% join$country)

open_space_pop_avg <- open_space_pop_avg %>% 
  filter(country %in% join$country)

transit_avg <- transit_avg %>% 
  filter(country %in% join$country)
```

There are a few data points with really high GDP.
```{r}
# filter three data frames by GDP
green_areas_avg <- green_areas_avg %>% 
  filter(gdp < 10000000000000)

transit_avg <- transit_avg %>% 
  filter(gdp < 10000000000000)

open_space_pop_avg <- open_space_pop_avg %>% 
  filter(gdp < 10000000000000)
```

## Prepare theme elements
```{r}
# save palette colors
mediumgreen <-  "#606C38"
darkgreen <- "#283618"
bg <- "#FEFAE0"
mediumorange <- "#DDA15E"
darkorange <- "#BC6C25"

# load font
font_add_google(name = "Montserrat", family = "mont")
font_add_google(name = "Google Sans", family = "google")
```

```{r}
# create ggplot theme to apply to all 3 plots
my_theme <- function(base_size = 17){
  theme_minimal(base_size = base_size) +
    theme(
    panel.grid = element_line(color = mediumorange, linewidth = 0.2),
    plot.title = element_text(family = "mont",
                                  size = rel(0.99),
                                  margin = margin(b = 7)),
    plot.background = element_rect(fill = bg,
                                   color = bg),
    panel.background = element_rect(fill = bg, 
                                    color = bg),
        plot.subtitle = element_text(family = "google", size = rel(0.70)),
        axis.text.y = element_text(hjust = 0, family = "google", size = rel(0.5), color = darkorange),
        axis.text.x = element_text(family = "google", size = rel(0.5), color = darkorange),
        axis.title = element_text(family = "google", size = rel(0.6), color = darkorange),
    )
}
```


## Plot

```{r}
#| warning: false
#| message: false
showtext_auto(enable = TRUE)

green <- ggplot(green_areas_avg, aes(avg_green_area, gdp)) +
  geom_point(alpha = 0.5,
             color = mediumgreen,
             fill = mediumgreen) +
  geom_smooth(method = "lm", # calculate linear relationship
              se = FALSE, # remove confidence interval
              linetype = 2, # dotted line
              linewidth = 1,
              color = darkgreen) + 
  labs(x = "Average % Share\nof Green Area",
       y = "National GDP (USD 2020)") +
  
  # adjust y-axis labels to read as "trillion"
  # set y limits to start at 0 -- geom_smooth goes below 0, throwing off axis
  scale_y_continuous( labels = scales::label_number(scale = 1e-12, suffix = " trillion"),
                      limits = c(0, NA)) +
  my_theme()

open <- ggplot(open_space_pop_avg, aes(avg_open_pop, gdp)) +
  geom_point(alpha = 0.5,
             color = mediumgreen,
             fill = mediumgreen) +
  geom_smooth(method = "lm",
              se = FALSE,
              linetype = 2, # dotted line
              linewidth = 1,
              color = darkgreen) +
   labs(x = "Average % of Population\nwith Access to Open Area",
       y = "National GDP (USD 2020)") +
  scale_y_continuous(labels = scales::label_number(scale = 1e-12, suffix = " trillion"),
                     limits = c(0, NA)) +
  my_theme()

transit <- ggplot(transit_avg, aes(avg_transport_pop, gdp)) +
  geom_point(alpha = 0.5,
             color = mediumgreen,
             fill = mediumgreen) +
  geom_smooth(method = "lm",
              se = FALSE,
              linetype = 2, # dotted line
              linewidth = 1,
              color = darkgreen) +
   labs(x = "Average % of Population\nwith Access to Public Transit",
       y = "National GDP (USD 2020)") +
   scale_y_continuous(labels = scales::label_number(scale = 1e-12, suffix = " trillion"),
                      limits = c(0, NA))+
  my_theme()

green + open + transit +
  plot_layout(axes="collect") + 
  plot_annotation(title = "Urban Resources & GDP: 111 countries worldwide",
                  subtitle = "Greater economic prosperity correlates with greater resources, such as\ngreen area, open space, and public transportation.",
                  caption = 'Access is defined as "Share of urban population with convenient access to public transport" by the United Nations.',
                  theme = theme(plot.background = element_rect(fill  = bg),
                                plot.title = element_text(family = "mont",
                                                          face = "bold",
                                  size = 17,
                                  margin = margin(b = 7, r = 7),
                                  color = darkgreen),
                                plot.subtitle = element_text(family = "mont",
                                  size = 12,
                                  margin = margin(b = 7),
                                  color = darkgreen),
                                plot.caption = element_text(family = "mont",
                                  size = 8,
                                 # margin = margin(b = 7),
                                  color = darkgreen)))

showtext_auto(enable = FALSE)
```


## Closing the connection!!
```{r}
DBI::dbDisconnect(conn, shutdown = TRUE)
```


