---
title: "wrangling"
format: html
---

```{r}
#| message: false
#| warning: false
# Load necessary libraries
library(tidyverse)
library(readxl)
library(here)
library(janitor)
```

### Read in data

```{r}
# share of urban population with convenient access to public transportation
transport_access <- read_excel(here("data", "transport_access.xlsx")) %>% 
  clean_names()

# open space for public use and its accessibility
public_space <- read_excel(here("data", "public_space.xlsx")) %>% 
  clean_names()

# average share of green space in urban area, over time
green_areas <- read_excel(here("data", "green_areas.xlsx"), sheet = 2) %>% 
  clean_names()
```

## Table Creation

### Location Table

All three datasets have the same few columns describing the location of their observation:
- county or territory code
- country or territory name
- SDG region
- SDG subregion
- city code
- city name

In the creation of the table, I will select these columns from one of the datasets and ensure that all data types are correct. `city_code` will be the primary key of this table.

```{r}
# select variables of interest
location <- public_space %>% 
  select(country_or_territory_code, country_or_territory_name,
         sdg_region, sdg_sub_region,
         city_code, city_name)

glimpse(location) # data types look good!
```
### Urban Resources (Transport Access and Public Space)

The transportation access and public spaces dataframes will be combined into one, representing urban resources. `city_code` will serve as the primary and foreign key in this table, referencing the location table.

```{r}
# select columns of interest: city_code, share of pop with public transport access, and data reference year
transport_access <- transport_access %>% 
  select(city_code, share_of_urban_population_with_convenient_access_to_public_transport_percent, data_reference_year) %>%
  
  # rename for clarity
  rename(transport_pop = share_of_urban_population_with_convenient_access_to_public_transport_percent)

# select columns of interest: city_code, share of city AREA that is open space, share of city POP that has access to public space, 
# and data reference year
public_space <- public_space %>% 
  select(city_code, average_share_of_the_built_up_area_of_cities_that_is_open_space_for_public_use_for_all_percent_a,
         average_share_of_urban_population_with_convenient_access_to_open_public_spaces_percent_b, data_reference_year) %>% 
  rename(open_space_area_share = average_share_of_the_built_up_area_of_cities_that_is_open_space_for_public_use_for_all_percent_a,
         open_space_pop = average_share_of_urban_population_with_convenient_access_to_open_public_spaces_percent_b)
```

### Green Areas

The green areas dataframe is currently in wide format.

```{r}
green_areas <- green_areas %>% 
  
  # select columns of interest
  select(city_code, average_share_of_green_area_in_city_urban_area_1990_percent, average_share_of_green_area_in_city_urban_area_2000_percent, average_share_of_green_area_in_city_urban_area_2010_percent, average_share_of_green_area_in_city_urban_area_2020_percent) %>% 
  
  # pivot longer
  pivot_longer(cols = starts_with("average"),
               names_to = "year",
               values_to = "green_area_share") %>% 
  
  # change observation names to just be years
  mutate(year = case_when(
    year == "average_share_of_green_area_in_city_urban_area_1990_percent" ~ "1990",
    year == "average_share_of_green_area_in_city_urban_area_2000_percent" ~ "2000",
    year == "average_share_of_green_area_in_city_urban_area_2010_percent" ~ "2010",
    year == "average_share_of_green_area_in_city_urban_area_2020_percent" ~ "2020"),
    
    # recode year as a factor (since we most likley wil be grouping by it)
    year = as.factor(year))
```

```{r}
glimpse(green_areas)
```
### Country and City Tables
```{r}
# select columns in city df
city <- location %>% 
  select(city_code, city_name, country_or_territory_code)

# select columns in country df
country <- location %>% 
  select(country_or_territory_code, country_or_territory_name, sdg_region, sdg_sub_region)

rm(location)
```

#### Add gdp data to country table.

```{r}
# read in 
gdp <- read_excel(here("data", "gdp.xls"), skip = 3) %>% 
  #clean_names() %>% 
  
  # keep only most recent GDPs
  select('Country Name', 'Country Code', '2023', '2024')

gdp <- gdp %>% 
  clean_names()
```

If GDP is missing from 2024, we replace it with the 2023 GDP. 

```{r}
gdp <- gdp %>% 
  
  # create a GDP column where...
  mutate(gdp = case_when(
    
    # ... if GDP from 2024 is missing, replace it with 2023
    is.na(x2024) ~ x2023,
    
    # otherwise, keep the 2024 value
    TRUE ~ x2024
  ))
```


The country codes between the World Bank data and United Nations data are not the same, so we will join GDP to our country database on country name. This will result in some missing countries.

```{r}
# join
country <- country %>% 
  left_join(gdp, by = join_by(country_or_territory_name == country_name))
```

```{r}
# how many countries are missing gdp?
country %>% 
  filter(is.na(as.numeric(gdp))) %>% 
  group_by(country_or_territory_name) %>% 
  summarize(count = n())
```
There are 44 observations that are in the UN data that are not represented in the World Bank data. However, some of those are territories (like "Central and Southern Asia"), whereas others are just named different things in the two datasets. We will standardize some of those names, and rejoin.

```{r}
# remove redundant names
country <- country %>% 
  filter(!country_or_territory_name %in% c("Australia and New Zealand", "Central and Southern Asia",
                                           "ECA", "ECE", "ECLAC", "ESCAP", "ESCWA", "Eastern and South-eastern Asia",
                                           "Latin America and the Caribbean", "Northern America and Europe", 
                                           "Oceania excluding Australia and New Zealand", "Puerto Rico",
                                           "Western Asia and Northern Africa"))

# how many countries are missing gdp now?
country %>% 
  filter(is.na(as.numeric(gdp))) %>% 
  group_by(country_or_territory_name) %>% 
  summarize(count = n())
```

We have decreased the number from 42 to 32. Now we can rename some of the countries where names are mismatched, and drop the rest.
```{r}
gdp <- gdp %>% 
  mutate(country_name = case_when(
    country_name == "United States" ~ "United States of America",
    country_name == "United Kingdom" ~ "United Kingdom of Great Britain and Northern Ireland",
    country_name == "Bolivia" ~ "Bolivia (Plurinational State of)",
    country_name == "Congo, Rep." ~ "Congo",
    country_name ==  "Congo, Dem. Rep" ~ "Democratic Republic of the Congo",
    country_name == "Bahamas, The" ~ "Bahamas",
    country_name == "Cote d'Ivoire" ~ "Côte d’Ivoire",
    country_name == "Egypt, Arab Rep." ~ "Egypt",
    country_name == "Gambia, The" ~ "Gambia",
    country_name == "Egypt, Arab Rep." ~ "Egypt",
    country_name == "Iran, Islamic Rep." ~ "Iran (Islamic Republic of)",
    country_name == "Korea, Dem. People's Rep." ~ "Democratic People's Republic of Korea",
    country_name == "Kyrgyz Republic" ~ "Kyrgyzstan",
    country_name == "Lao PDR" ~ "Lao People's Democratic Republic",
    country_name == "Moldova" ~ "Republic of Moldova",
    country_name == "Slovak Republic" ~ "Slovakia",
    country_name == "Somalia, Fed. Rep." ~ "Somalia",
    country_name == "St. Kitts and Nevis" ~ "Saint Kitts and Nevis",
    country_name == "St. Lucia" ~ "Saint Lucia",
    country_name == "St. Vincent and the Grenadines" ~ "Saint Vincent and the Grenadines",
    country_name == "Tanzania" ~ "United Republic of Tanzania",
    country_name == "Turkiye" ~ "Türkiye",
    country_name == "Venezuela, RB" ~ "Venezuela (Bolivarian Republic of)",
    country_name == "West Bank and Gaza" ~ "State of Palestine",
    country_name == "Korea, Rep." ~ "Republic of Korea",
    country_name == "Congo, Dem. Rep." ~ "Democratic Republic of the Congo",
    country_name == "Netherlands" ~ "Netherlands (Kingdom of the)",
    country_name == "Yemen, Rep." ~ "Yemen",
  TRUE ~ country_name
))

# rejoin
country <- country %>% 
  left_join(gdp, by = join_by(country_or_territory_name == country_name))
```

```{r}
# how many countries are missing gdp now?
country %>% 
  filter(is.na(as.numeric(gdp.y))) %>% 
  group_by(country_or_territory_name) %>% 
  summarize(count = n())
```

These are all counties where GDP from World Bank is actually NA (and not due to differences in country name).

```{r}
country <- country %>% 
  select(country_or_territory_code, country_or_territory_name, sdg_region, sdg_sub_region, gdp.y) %>% 
  rename(annual_gdp = gdp.y)

rm(gdp)
```

## Finl Check
```{r}
glimpse(city)
```
```{r}
glimpse(country)
```
```{r}
glimpse(green_areas)
```
```{r}
public_space <- public_space %>% 
  mutate(reference_year = as.factor(data_reference_year)) %>% 
  select(-data_reference_year)

glimpse(public_space)
```
```{r}
transport_access <- transport_access %>% 
  mutate(reference_year = as.factor(data_reference_year)) %>% 
  select(-data_reference_year)

glimpse(transport_access)
```
Fixing duplicate issue in city
```{r}
city <- city %>% 
  filter(!duplicated(city_code))
```


## Saving
```{r}
write_csv(city, file.path("data", "city.csv"))
write_csv(country, file.path("data", "country.csv"))
write_csv(green_areas, file.path("data", "green_areas.csv"))
write_csv(public_space, file.path("data", "public_space.csv"))
write_csv(transport_access, file.path("data", "transport_access.csv"))
```

