---
title: "UN-Habitat Urban Indicators and World Bank GDP data wrangling"
author: "Sofia Sarak"
format: html
editor: 
  markdown: 
    wrap: 72
---

The following document contains the necessary wrangling and cleaning
necessary to upload the data into SQL and create five distinct tables:

-   Country (with GDP data)

-   Open_space

-   Green_Areas

-   City

-   Transport

Raw data can be downloaded from the links in the GitHub repository.

```{r}
#| message: false
#| warning: false
# Load necessary libraries
library(tidyverse)
library(readxl)
library(here)
library(janitor)
```

### Read in UN-Habitat Urban Indicators data

-   <div>

    ```{r}
    # share of urban population with convenient access to public transportation
    transport_access <- read_excel(here("data", "transport_access.xlsx")) %>% 
      clean_names()

    # open space for public use and its accessibility
    public_space <- read_excel(here("data", "public_space.xlsx")) %>% 
      clean_names()

    # average share of green space in urban area, over time
    green_areas <- read_excel(here("data", "green_areas.xlsx"), sheet = 2) %>% 
      clean_names()
    ```

    </div>

## Table Creation

### Urban Resources (Transport Access and Public Space)

The transportation access and public spaces dataframes will be combined
into one, representing urban resources. `city_code` will serve as the
primary and foreign key in this table, referencing the location table.

```{r}
# select columns of interest: city_code, share of pop with public transport access, and data reference year
transport_access <- transport_access %>% 
  select(city_code, share_of_urban_population_with_convenient_access_to_public_transport_percent, data_reference_year) %>%
  
  # rename for clarity
  rename(transport_pop = share_of_urban_population_with_convenient_access_to_public_transport_percent)

# select columns of interest: city_code, share of city AREA that is open space, share of city POP that has access to public space, 
# and data reference year
public_space <- public_space %>% 
  select(city_code, average_share_of_the_built_up_area_of_cities_that_is_open_space_for_public_use_for_all_percent_a,
         average_share_of_urban_population_with_convenient_access_to_open_public_spaces_percent_b, data_reference_year) %>% 
  rename(open_space_area_share = average_share_of_the_built_up_area_of_cities_that_is_open_space_for_public_use_for_all_percent_a,
         open_space_pop = average_share_of_urban_population_with_convenient_access_to_open_public_spaces_percent_b)
```

### Green Areas

The green areas dataframe is currently in wide format.

```{r}
green_areas <- green_areas %>% 
  
  # select columns of interest
  select(city_code, average_share_of_green_area_in_city_urban_area_1990_percent, average_share_of_green_area_in_city_urban_area_2000_percent, average_share_of_green_area_in_city_urban_area_2010_percent, average_share_of_green_area_in_city_urban_area_2020_percent) %>% 
  
  # pivot longer
  pivot_longer(cols = starts_with("average"),
               names_to = "year",
               values_to = "green_area_share") %>% 
  
  # change observation names to just be years
  mutate(year = case_when(
    year == "average_share_of_green_area_in_city_urban_area_1990_percent" ~ "1990",
    year == "average_share_of_green_area_in_city_urban_area_2000_percent" ~ "2000",
    year == "average_share_of_green_area_in_city_urban_area_2010_percent" ~ "2010",
    year == "average_share_of_green_area_in_city_urban_area_2020_percent" ~ "2020"),
    
    # recode year as a factor (since we most likley wil be grouping by it)
    year = as.factor(year))
```

```{r}
glimpse(green_areas)
```

### Location Table

All three datasets have the same few columns describing the location of
their observation:

-   county or territory code

-   country or territory name

-   SDG region

-   SDG subregion

-   city code

-   city name

In the creation of the table, I will select these columns from one of
the datasets and ensure that all data types are correct. `city_code`
will be the primary key of this table, and it will house information on
each city's country and region.

```{r}
# select variables of interest
location <- public_space %>% 
  select(country_or_territory_code, country_or_territory_name,
         sdg_region, sdg_sub_region,
         city_code, city_name)

glimpse(location) # data types look good!
```

### Country and City Tables

```{r}
# select columns in city df
city <- location %>% 
  select(city_code, city_name, country_or_territory_code)

# select columns in country df
country <- location %>% 
  select(country_or_territory_code, country_or_territory_name, sdg_region, sdg_sub_region)

rm(location)
```

#### Add gdp data to country table.

```{r}
# read in 
gdp <- read_excel(here("data", "gdp.xls"), skip = 3) %>% 
  #clean_names() %>% 
  
  # keep only most recent GDPs
  select('Country Name', 'Country Code', '2023', '2024')

gdp <- gdp %>% 
  clean_names()
```

If GDP is missing from 2024, we replace it with the 2023 GDP.

```{r}
gdp <- gdp %>% 
  
  # create a GDP column where...
  mutate(gdp = case_when(
    
    # ... if GDP from 2024 is missing, replace it with 2023
    is.na(x2024) ~ x2023,
    
    # otherwise, keep the 2024 value
    TRUE ~ x2024
  ))
```

The country codes between the World Bank data and United Nations data
are not the same, so we will join GDP to our country database on country
name. This will result in some missing countries.

```{r}
# join
country <- country %>% 
  left_join(gdp, by = join_by(country_or_territory_name == country_name))
```

```{r}
# how many countries are missing gdp?
country %>% 
  filter(is.na(as.numeric(gdp))) %>% 
  group_by(country_or_territory_name) %>% 
  summarize(count = n())
```

There are 44 observations that are in the UN data that are not
represented in the World Bank data. However, some of those are
territories (like "Central and Southern Asia"), whereas others are just
named different things in the two datasets. We will standardize some of
those names, and rejoin.

```{r}
# remove redundant names
country <- country %>% 
  filter(!country_or_territory_name %in% c("Australia and New Zealand", "Central and Southern Asia",
                                           "ECA", "ECE", "ECLAC", "ESCAP", "ESCWA", "Eastern and South-eastern Asia",
                                           "Latin America and the Caribbean", "Northern America and Europe", 
                                           "Oceania excluding Australia and New Zealand", "Puerto Rico",
                                           "Western Asia and Northern Africa"))

# how many countries are missing gdp now?
country %>% 
  filter(is.na(as.numeric(gdp))) %>% 
  group_by(country_or_territory_name) %>% 
  summarize(count = n())
```

We have decreased the number from 42 to 32. Now we can rename some of
the countries where names are mismatched, and drop the rest.

```{r}
gdp <- gdp %>% 
  mutate(country_name = case_when(
    country_name == "United States" ~ "United States of America",
    country_name == "United Kingdom" ~ "United Kingdom of Great Britain and Northern Ireland",
    country_name == "Bolivia" ~ "Bolivia (Plurinational State of)",
    country_name == "Congo, Rep." ~ "Congo",
    country_name ==  "Congo, Dem. Rep" ~ "Democratic Republic of the Congo",
    country_name == "Bahamas, The" ~ "Bahamas",
    country_name == "Cote d'Ivoire" ~ "Côte d’Ivoire",
    country_name == "Egypt, Arab Rep." ~ "Egypt",
    country_name == "Gambia, The" ~ "Gambia",
    country_name == "Egypt, Arab Rep." ~ "Egypt",
    country_name == "Iran, Islamic Rep." ~ "Iran (Islamic Republic of)",
    country_name == "Korea, Dem. People's Rep." ~ "Democratic People's Republic of Korea",
    country_name == "Kyrgyz Republic" ~ "Kyrgyzstan",
    country_name == "Lao PDR" ~ "Lao People's Democratic Republic",
    country_name == "Moldova" ~ "Republic of Moldova",
    country_name == "Slovak Republic" ~ "Slovakia",
    country_name == "Somalia, Fed. Rep." ~ "Somalia",
    country_name == "St. Kitts and Nevis" ~ "Saint Kitts and Nevis",
    country_name == "St. Lucia" ~ "Saint Lucia",
    country_name == "St. Vincent and the Grenadines" ~ "Saint Vincent and the Grenadines",
    country_name == "Tanzania" ~ "United Republic of Tanzania",
    country_name == "Turkiye" ~ "Türkiye",
    country_name == "Venezuela, RB" ~ "Venezuela (Bolivarian Republic of)",
    country_name == "West Bank and Gaza" ~ "State of Palestine",
    country_name == "Korea, Rep." ~ "Republic of Korea",
    country_name == "Congo, Dem. Rep." ~ "Democratic Republic of the Congo",
    country_name == "Netherlands" ~ "Netherlands (Kingdom of the)",
    country_name == "Yemen, Rep." ~ "Yemen",
  TRUE ~ country_name
))

# rejoin
country <- country %>% 
  left_join(gdp, by = join_by(country_or_territory_name == country_name))
```

```{r}
# how many countries are missing gdp now?
country %>% 
  filter(is.na(as.numeric(gdp.y))) %>% 
  group_by(country_or_territory_name) %>% 
  summarize(count = n())
```

These are all counties where GDP from World Bank is actually NA (and not
due to differences in country name).

```{r}
country <- country %>% 
  select(country_or_territory_code, country_or_territory_name, sdg_region, sdg_sub_region, gdp.y) %>% 
  rename(annual_gdp = gdp.y)

rm(gdp)
```

## Addressing other issues

After trying to upload these versions of the data into SQL, a few issues
cropped up, including: - duplicates - primary keys not actually being
unique - null values

The following code cleans up these problems.

```{r}
city <-  city |>
  filter(!duplicated(city_code))

## Country -- drop observations where gdp is NA
# how many countries are missing gdp?
missing <-  country %>% 
       filter(is.na(annual_gdp))
unique(missing$country_or_territory_code)
# only 8! we can drop them (but will keep this in mind during our analysis)

country <- country %>% 
  filter(!is.na(annual_gdp))

# group by country code --> currently is repeated a bunch
country <- country %>% 
  filter(!duplicated(country_or_territory_code))

## Green spaces -- explore city_code duplicates
green %>% 
  filter(duplicated(city_code))

# remove NAs
green <- green %>% 
  filter(!is.na(city_code))

# Because our data is in long format, all city_codes repeat 3 times (one for each year). In SQL, we will
# set the primary key to be a compound of city_code and year


## Public Space -- explore NAs and year column
# how many NA values are there?
sum(is.na(space$open_space_area_share))
sum(is.na(space$open_space_pop))

# what do those rows looks like?
filter(space, is.na(space$open_space_pop))

# it looks like sometimes there is population data and not area data, and vice versa. I don't want to drop valuable info
# from one column just because the other is missing, so will remove NOT NULL from the SQL

# there were some issues with the year column as well --> warning that is is of class BIGINT?
unique(space$reference_year)
glimpse(space)

# drop NAs in city_code column
space <- space %>% 
  filter(!is.na(city_code))

# multiple cities with data from diff years: keep most recent
space_max <- space %>% 
  group_by(city_code) %>% 
  summarize(max_year = max(reference_year))

space <- space %>% 
  left_join(space_max) %>% 
  filter(reference_year == max_year) %>% 
  filter(!is.na(city_code)) %>% 
  select(!max_year)

# some rows are duplicated where one vars is NA and the other is not
# here, I combine those values so each duplicate has the same values (both pop and area)
space <-space %>% 
  group_by(city_code) %>% 
  mutate(open_space_area_share = sum(open_space_area_share, na.rm = TRUE),
         open_space_pop = sum(open_space_pop, na.rm = TRUE))

# now, I'll drop the duplictaes
space <- space %>% 
  filter(!duplicated(city_code))

## Transport Access -- explore duplicate city_codes
dup <- transport %>% 
  filter(duplicated(city_code))
# looks like some observations have multiple reference years. This only occurs a handful of times, so to keep data consistent
# I will keep the data from the most recent reference year. There are also a few NA city_codes, that I will drop.

# find max reference year of duplicates
dup_max <- dup %>% 
  group_by(city_code) %>% 
  summarize(max_year = max(reference_year))

# attach max year to each city code
dup <- dup %>% 
  left_join(dup_max)

# keep only transport_pop from the max_year
dup <- dup %>% 
  filter(reference_year == max_year)

# now that we know the workflow works, apply it to all of transport
transport_max <- transport %>% 
  group_by(city_code) %>% 
  summarize(max_year = max(reference_year))

transport <- transport %>% 
  left_join(transport_max) %>% 
  filter(reference_year == max_year) %>% 
  filter(!is.na(city_code)) %>% 
  select(!max_year)

# confirm there are no duplicates anymore
transport %>% 
  filter(duplicated(city_code))
```

## Final Check

```{r}
glimpse(city)
```

```{r}
glimpse(country)
```

```{r}
glimpse(green_areas)
```

```{r}
public_space <- public_space %>% 
  mutate(reference_year = as.factor(data_reference_year)) %>% 
  select(-data_reference_year)

glimpse(public_space)
```

```{r}
transport_access <- transport_access %>% 
  mutate(reference_year = as.factor(data_reference_year)) %>% 
  select(-data_reference_year)

glimpse(transport_access)
```

## Saving

```{r}
write_csv(city, file.path("data", "city.csv"))
write_csv(country, file.path("data", "country.csv"))
write_csv(green_areas, file.path("data", "green_areas.csv"))
write_csv(public_space, file.path("data", "public_space.csv"))
write_csv(transport_access, file.path("data", "transport_access.csv"))
```
