---
title: "wrangling"
format: html
---

```{r}
#| message: false
#| warning: false
# Load necessary libraries
library(tidyverse)
library(readxl)
library(here)
library(janitor)
```

### Read in data

```{r}
# share of urban population with convenient access to public transportation
transport_access <- read_excel(here("data", "transport_access.xlsx")) %>% 
  clean_names()

# open space for public use and its accessibility
public_space <- read_excel(here("data", "public_space.xlsx")) %>% 
  clean_names()

# average share of green space in urban area, over time
green_areas <- read_excel(here("data", "green_areas.xlsx"), sheet = 2) %>% 
  clean_names()
```

## Table Creation

### Location Table

All three datasets have the same few columns describing the location of their observation:
- county or territory code
- country or territory name
- SDG region
- SDG subregion
- city code
- city name

In the creation of the table, I will select these columns from one of the datasets and ensure that all data types are correct. `city_code` will be the primary key of this table.

```{r}
# select variables of interest
location <- public_space %>% 
  select(country_or_territory_code, country_or_territory_name,
         sdg_region, sdg_sub_region,
         city_code, city_name)

glimpse(location) # data types look good!
```
### Urban Resources (Transport Access and Public Space)

The transportation access and public spaces dataframes will be combined into one, representing urban resources. `city_code` will serve as the primary and foreign key in this table, referencing the location table.

```{r}
# select columns of interest: city_code, share of pop with public transport access, and data reference year
transport_access <- transport_access %>% 
  select(city_code, share_of_urban_population_with_convenient_access_to_public_transport_percent, data_reference_year) %>%
  
  # rename for clarity
  rename(transport_pop = share_of_urban_population_with_convenient_access_to_public_transport_percent)

# select columns of interest: city_code, share of city AREA that is open space, share of city POP that has access to public space, 
# and data reference year
public_space <- public_space %>% 
  select(city_code, average_share_of_the_built_up_area_of_cities_that_is_open_space_for_public_use_for_all_percent_a,
         average_share_of_urban_population_with_convenient_access_to_open_public_spaces_percent_b, data_reference_year) %>% 
  rename(open_space_area = average_share_of_the_built_up_area_of_cities_that_is_open_space_for_public_use_for_all_percent_a,
         open_space_pop = average_share_of_urban_population_with_convenient_access_to_open_public_spaces_percent_b)
```

### Green Areas

The green areas dataframe is currently in wide format.

```{r}
green_areas <- green_areas %>% 
  
  # select columns of interest
  select(city_code, average_share_of_green_area_in_city_urban_area_1990_percent, average_share_of_green_area_in_city_urban_area_2000_percent, average_share_of_green_area_in_city_urban_area_2010_percent, average_share_of_green_area_in_city_urban_area_2020_percent) %>% 
  
  # pivot longer
  pivot_longer(cols = starts_with("average"),
               names_to = "year",
               values_to = "city_green_area_share") %>% 
  
  # change observation names to just be years
  mutate(year = case_when(
    year == "average_share_of_green_area_in_city_urban_area_1990_percent" ~ "1990",
    year == "average_share_of_green_area_in_city_urban_area_2000_percent" ~ "2000",
    year == "average_share_of_green_area_in_city_urban_area_2010_percent" ~ "2010",
    year == "average_share_of_green_area_in_city_urban_area_2020_percent" ~ "2020"),
    
    # recode year as a factor (since we most likley wil be grouping by it)
    year = as.factor(year))
```

```{r}
glimpse(green_areas)
```
### Country and City Tables
```{r}
# select columns in city df
city <- location %>% 
  select(city_code, city_name, country_or_territory_code)

# select columns in country df
country <- location %>% 
  select(country_or_territory_code, country_or_territory_name, sdg_region, sdg_sub_region)

rm(location)
```

#### Add gdp data to country table.

```{r}
# read in 
gdp <- read_excel(here("data", "gdp.xls"), skip = 3) %>% 
  #clean_names() %>% 
  
  # keep only most recent GDPs
  select('Country Name', 'Country Code', '2023', '2024')

gdp <- gdp %>% 
  clean_names()
```

If GDP is missing from 2024, we replace it with the 2023 GDP. 

```{r}

```

