---
title: "wrangling"
format: html
---

```{r}
#| message: false
#| warning: false
# Load necessary libraries
library(tidyverse)
library(readxl)
library(here)
library(janitor)
```

### Read in data

```{r}
# share of urban population with convenient access to public transportation
transport_access <- read_excel(here("data", "SDG_11-2-1.xlsx")) %>% 
  clean_names()

# open space for public use and its accessibility
public_space <- read_excel(here("data", "SDG_11-7-1.xlsx")) %>% 
  clean_names()

# average share of green space in urban area, over time
green_areas <- read_excel(here("data", "green_areas.xlsx"), sheet = 2) %>% 
  clean_names()
```

## Table Creation

### Location Table

All three datasets have the same few columns describing the location of their observation:
- county or territory code
- country or territory name
- SDG region
- SDG subregion
- city code
- city name

In the creation of the table, I will select these columns from one of the datasets and ensure that all data types are correct. `city_code` will be the primary key of this table.

```{r}
# select variables of interest
location <- public_space %>% 
  select(country_or_territory_code, country_or_territory_name,
         sdg_region, sdg_sub_region,
         city_code, city_name)

glimpse(location) # data types look good!
```
### Urban Resources Table

The transportation access and public spaces dataframes will be combined into one, representing urban resources. `city_code` will serve as the primary and foreign key in this table, referencing the location table.

```{r}
transport_columns <- transport_access %>% 
  select(city_code, share_of_urban_population_with_convenient_access_to_public_transport_percent) %>% 
  rename(transport_access = share_of_urban_population_with_convenient_access_to_public_transport_percent)

public_space_columns <- public_space %>% 
  select(city_code, average_share_of_the_built_up_area_of_cities_that_is_open_space_for_public_use_for_all_percent_a,
         average_share_of_urban_population_with_convenient_access_to_open_public_spaces_percent_b) %>% 
  rename(open_space = average_share_of_the_built_up_area_of_cities_that_is_open_space_for_public_use_for_all_percent_a,
         access = average_share_of_urban_population_with_convenient_access_to_open_public_spaces_percent_b)
```

